Cultural Evolutionary Dynamics with Success and Influence biases

### 2022-09-14 ms/PrestigeBias.tex
update

This bias depends on the state of a role-model rather than on its trait, in contrast to frequency biases such as conformity \citep{kin_selection}, % TODO are you sure this is the correct reference? maybe change to Denton et al?
which depend on the frequency of a trait in the population or in a sample of role-models. 
 We define a model of cultural transmission with prestige bias that combines both success and influence biases, provide analytic approximations for this model, and analyze its dynamics.

@@

\begin{proof}
The initial prestige of role-model $j$ based on \cref{eq:prestige} is
\begin{equation}\label{eq:initial_prob}
G_{1,j} = \frac{\alpha_j\cdot\beta(A_j+e)}{\sum\limits_{m=1}^{N} \alpha_m\cdot\beta(A_m+e)} \;.
\end{equation}
The denominator of \cref{eq:initial_prob} can also be formulated as:
\begin{equation}\label{eq:denominator}
 \sum\limits_{m=1}^{N}\alpha_m\beta(A_m+e) = N \cdot \overline{\alpha \cdot \beta(A+e)} \;,
\end{equation}
where $\overline{\alpha\beta(A+e)}$ is the mean value of $\alpha_m\cdot\beta(A_m+e)$ for all $m$.
Using \cref{eq:denominator} and \textbf{Corollary 1} we get,
\begin{equation}
\E[K_{Nj}] = \alpha_j \cdot \beta(A_j+e) \bigg/ \overline{\alpha \cdot \beta(A+e)} \;,
\end{equation}
completing the proof.
\end{proof}

@@

\begin{proof}
Let $X$ be the number of individuals of type $\hat{A}$ such that $x=X/N$. $X'$ is the number of individuals with $\hat{A}$ in the next generation.
The expected number of individuals is (due to the DM approximation),
\begin{equation}
E[X'] = N  \frac{\alpha_1}{\alpha_1+\alpha_2} \;,
\end{equation}
where $\alpha_1 = \alpha' X$ and $\alpha_2 = \alpha'(N-X)\beta$, from  \cref{eq:binary-model}.
To use frequencies instead of counts, $E[x'] = E[X'/N] = \frac{1}{N}E[X']$.
Putting it together,
\begin{equation}
\begin{split}
E[x'] &= \frac{1}{N}N\frac{\alpha' xN}{\alpha' xN + \alpha' N (1-x)\beta}
	  = \frac{x}{x + (1-x)\beta} \\
	  &= \frac{x}{x + (1-x) -(1-x) + (1-x)\beta}
	  = x \frac{1}{1 -(1-x)(1-\beta)}  \\
	  &= x \big(1 + (1-x)(1-\beta) + o(\beta)\big)
	  = x + x(1-x)(1-\beta) + o(\beta) \;,
\end{split}
\end{equation}
following \citet[p.~253, ch~7.2]{durret} and because $1/(1-y)=1+y+y^2+\ldots$.

Following \citet[ch~7.2]{durret}, we assume $\beta \approx 1$, such that
\begin{equation}
\frac{x}{x + (1-x)\beta} \approx x \,
\end{equation}
and for the entire variance expression we get
\begin{equation}
%\begin{split}
V(x') \approx  \frac{1}{N} x(1-x)
\Big(\frac{N + \alpha' xN + \alpha' N - \alpha' xN}{1 + \alpha' xN + \alpha' N - \alpha' xN}\Big)
= x(1-x)\Big(\frac{1 + \alpha'}{1 + \alpha' N}\Big) \;.
%\end{split}
\end{equation}
Now because $x$ is a constant, $V(x) = 0$, and therefore
\begin{equation}\label{eq:var_diff_durret}
V(x'-x) = V(x') - V(x) \approx  x(1-x)(\frac{1 + \alpha'}{1 + \alpha' N}) \;.
\end{equation}
$\alpha'$ is the odds ratio of the bias weight, 
\begin{equation}\label{eq:success_ratio}
\alpha' = \frac{\alpha}{1-\alpha} \;.
\end{equation}
Combining \cref{eq:var_diff_durret} and \cref{eq:success_ratio} we get:
\begin{equation}\label{eq:const_var}
\begin{split}
V(x'-x) \approx x(1-x)\Big(\frac{1 + \frac{\alpha}{1-\alpha}}{1 + \frac{\alpha}{1-\alpha} N}\Big)
% &= x(1-x)(\frac{\frac{1-\alpha+\alpha}{1-\alpha}}{\frac{1-\alpha+\alpha N}{1-\alpha}})\\
% &= x(1-x)(\frac{1}{1- \alpha(1-N)})\\
  = x(1-x)(\frac{1}{\alpha N + (1-\alpha)}) \;.
\end{split}
\end{equation}
This gives the diffusion term of the diffusion equation, which completes the proof.
\end{proof}

@@

\begin{proof}
Let $s_t=N(1-\beta_t)$, and $S_n=\sum\limits_{i=1}^n s_i$, where $\beta_t$ is $\beta(A)$ at generation $t$.
We prove by induction both terms in \cref{eq:ch_expected_and_var}.
From \cref{eq:expec_freq} we know that
\begin{equation}\label{eq:ch_1}
%\begin{split}
E\left[\frac{X_{t+1}}{N}-\frac{X_t}{N}\bigg|X_t\right] 
= \frac{X_t}{N}\left(1-\frac{X_t}{N}\right)(1-\beta_{t+1}) 
= \frac{1}{N}\frac{X_t}{N}\left(1-\frac{X_t}{N}\right)s_{t+1} \;.
%\end{split}
\end{equation}
Also note that using the definition of $V(y)=E[y^2]-(E[y])^2$
\begin{equation}
\begin{split}
E\left[\frac{X_t}{N}\left(1-\frac{X_t}{N}\right)\right] 
&= E\left[\frac{X_t}{N}-\left(\frac{X_t}{N}\right)^2\right] 
= E\left[\frac{X_t}{N}\right] - E\left[\left(\frac{X_t}{N}\right)^2\right] \\
&= E\left[\frac{X_t}{N}\right] - V\left(\frac{X_t}{N}\right) - \left(E\left[\frac{X_t}{N}\right]\right)^2 \;.
\end{split}
\end{equation}

@@

We also analyzed a cyclic changing environment in which the identity of the success-biased trait switches after a fixed amount of generations~(\Cref{fig:ch_env_alpha_beta}).
We find that, similarly to the constant environment approximation, a change in the success-bias weight % TODO which symbol is that? was that it's name in the model section?
 has no negative effects on the goodness-of-fit of the approximation to simulation results.
We also showed that this approximation is more sensitive to changes in the success coefficient value % TODO which symbol is that? was that it's name in the model section?
than the constant environment approximation, and a lower value is required to have a good fit. The ratio between the number of generation in which the rare phenotype is under positive transmission bias and the number of generations in which it is under negative bias does not affect the goodness-of-fit of the approximation.

@@

\pagebreak
\end{document}

>>>>

This bias depends on the state of a role-model rather than on its trait, in contrast to frequency biases such as conformity \citep{conformity}, which depend on the frequency of a trait in the population or in a sample of role-models. 
 We define a model of cultural transmission with prestige bias that combines both success and influence biases, provide analytic approximations for this model, and analyze its dynamics.

@@

\begin{proof}\cref{append:GBD}
\end{proof}

@@

\begin{proof}
\cref{append:drift_diff_const}
\end{proof}

@@

\begin{proof}
\cref{append:drift_diff_chang}
\end{proof}

@@

We also analyzed a cyclic changing environment in which the identity of the success-biased trait switches after a fixed amount of generations~(\Cref{fig:ch_env_alpha_beta}).
We find that, similarly to the constant environment approximation, a change in the success-bias weight $\alpha$ has no negative effects on the goodness-of-fit of the approximation to simulation results.
We also showed that this approximation is more sensitive to changes in the success coefficient $\beta$ than the constant environment approximation, and a lower value is required to have a good fit. The ratio between the number of generation in which the rare phenotype is under positive transmission bias and the number of generations in which it is under negative bias does not affect the goodness-of-fit of the approximation.

@@

\appendix
\section{General Binomial Distribution approximation proof}\label{append:GBD}
\begin{proof}
The initial prestige of role-model $j$ based on \cref{eq:prestige} is
\begin{equation}\label{eq:initial_prob}
G_{1,j} = \frac{\alpha_j\cdot\beta(A_j+e)}{\sum\limits_{m=1}^{N} \alpha_m\cdot\beta(A_m+e)} \;.
\end{equation}
The denominator of \cref{eq:initial_prob} can also be formulated as:
\begin{equation}\label{eq:denominator}
 \sum\limits_{m=1}^{N}\alpha_m\beta(A_m+e) = N \cdot \overline{\alpha \cdot \beta(A+e)} \;,
\end{equation}
where $\overline{\alpha\beta(A+e)}$ is the mean value of $\alpha_m\cdot\beta(A_m+e)$ for all $m$.
Using \cref{eq:denominator} and \textbf{Corollary 1} we get,
\begin{equation}
\E[K_{Nj}] = \alpha_j \cdot \beta(A_j+e) \bigg/ \overline{\alpha \cdot \beta(A+e)} \;,
\end{equation}
\end{proof}

\section{Drift and diffusion in a constant environment proof}\label{append:drift_diff_const}
\begin{proof}
Let $X$ be the number of individuals of type $\hat{A}$ such that $x=X/N$. $X'$ is the number of individuals with $\hat{A}$ in the next generation.
The expected number of individuals is (due to the DM approximation),
\begin{equation}
E[X'] = N  \frac{\alpha_1}{\alpha_1+\alpha_2} \;,
\end{equation}
where $\alpha_1 = \alpha' X$ and $\alpha_2 = \alpha'(N-X)\beta$, from  \cref{eq:binary-model}.
To use frequencies instead of counts, $E[x'] = E[X'/N] = \frac{1}{N}E[X']$.
Putting it together,
\begin{equation}
\begin{split}
E[x'] &= \frac{1}{N}N\frac{\alpha' xN}{\alpha' xN + \alpha' N (1-x)\beta}
	  = \frac{x}{x + (1-x)\beta} \\
	  &= \frac{x}{x + (1-x) -(1-x) + (1-x)\beta}
	  = x \frac{1}{1 -(1-x)(1-\beta)}  \\
	  &= x \big(1 + (1-x)(1-\beta) + o(\beta)\big)
	  = x + x(1-x)(1-\beta) + o(\beta) \;,
\end{split}
\end{equation}
following \citet[p.~253, ch~7.2]{durret} and because $1/(1-y)=1+y+y^2+\ldots$.

Following \citet[ch~7.2]{durret}, we assume $\beta \approx 1$, such that
\begin{equation}
\frac{x}{x + (1-x)\beta} \approx x \,
\end{equation}
and for the entire variance expression we get
\begin{equation}
%\begin{split}
V(x') \approx  \frac{1}{N} x(1-x)
\Big(\frac{N + \alpha' xN + \alpha' N - \alpha' xN}{1 + \alpha' xN + \alpha' N - \alpha' xN}\Big)
= x(1-x)\Big(\frac{1 + \alpha'}{1 + \alpha' N}\Big) \;.
%\end{split}
\end{equation}
Now because $x$ is a constant, $V(x) = 0$, and therefore
\begin{equation}\label{eq:var_diff_durret}
V(x'-x) = V(x') - V(x) \approx  x(1-x)(\frac{1 + \alpha'}{1 + \alpha' N}) \;.
\end{equation}
$\alpha'$ is the odds ratio of the bias weight, 
\begin{equation}\label{eq:success_ratio}
\alpha' = \frac{\alpha}{1-\alpha} \;.
\end{equation}
Combining \cref{eq:var_diff_durret} and \cref{eq:success_ratio} we get:
\begin{equation}\label{eq:const_var}
\begin{split}
V(x'-x) \approx x(1-x)\Big(\frac{1 + \frac{\alpha}{1-\alpha}}{1 + \frac{\alpha}{1-\alpha} N}\Big)
% &= x(1-x)(\frac{\frac{1-\alpha+\alpha}{1-\alpha}}{\frac{1-\alpha+\alpha N}{1-\alpha}})\\
% &= x(1-x)(\frac{1}{1- \alpha(1-N)})\\
  = x(1-x)(\frac{1}{\alpha N + (1-\alpha)}) \;.
\end{split}
\end{equation}
This gives the diffusion term of the diffusion equation.
\end{proof}

\section{Drift and diffusion in a changing environment proof}\label{append:drift_diff_chang}
\begin{proof}
Let $s_t=N(1-\beta_t)$, and $S_n=\sum\limits_{i=1}^n s_i$, where $\beta_t$ is $\beta(A)$ at generation $t$.
We prove by induction both terms in \cref{eq:ch_expected_and_var}.
From \cref{eq:expec_freq} we know that
\begin{equation}\label{eq:ch_1}
%\begin{split}
E\left[\frac{X_{t+1}}{N}-\frac{X_t}{N}\bigg|X_t\right] 
= \frac{X_t}{N}\left(1-\frac{X_t}{N}\right)(1-\beta_{t+1}) 
= \frac{1}{N}\frac{X_t}{N}\left(1-\frac{X_t}{N}\right)s_{t+1} \;.
%\end{split}
\end{equation}
Also note that using the definition of $V(y)=E[y^2]-(E[y])^2$
\begin{equation}
\begin{split}
E\left[\frac{X_t}{N}\left(1-\frac{X_t}{N}\right)\right] 
&= E\left[\frac{X_t}{N}-\left(\frac{X_t}{N}\right)^2\right] 
= E\left[\frac{X_t}{N}\right] - E\left[\left(\frac{X_t}{N}\right)^2\right] \\
&= E\left[\frac{X_t}{N}\right] - V\left(\frac{X_t}{N}\right) - \left(E\left[\frac{X_t}{N}\right]\right)^2 \;.
\end{split}
\end{equation}

\end{document}
